---
title: "RNA-seq model selection"
subtitle: "South Africa: TB-induced RSTR vs LTBI"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---
# Background

The purpose of this workflow is to determine co-variates to include in DEG analysis.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
    # Multi-panel figures for ggplot
    library(cowplot)
library(venn)
library(biomaRt)
# Print tty table to knit file
library(knitr)
library(kableExtra)

`%notin%` <- Negate(`%in%`)
```

Set seed

```{r}
set.seed(4389)
```

# Customization

Set variable names and cutoffs for this workflow.

```{r}
#Rdata file WITHIN project directory that holds cleaned data
data.file <- "data_clean/RSTR_SA_dat_clean.RData"

#Prefix to give file names
basename <- "SA_RSTR.Mtb"
#Define variable(s) of interest
#Used in PCA plots and to select significant genes to be used in module building
vars_of_interest <- c("condition", "Sample_Group")
```

# Load data

```{r}
#Load data
load(data.file)
```

This includes in the following samples.

```{r echo=FALSE}
dat.abund.norm.voom$targets %>% 
  group_by_at(vars(vars_of_interest)) %>% 
  tally() %>% 
  
  kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

# Data exploration
## PCA
### Main variables

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Calculate PCA
PCA.all <- as.data.frame(dat.abund.norm.voom$E) %>% 
  t() %>% 
  prcomp()
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Get XY labels
PCA.labs <- list()
PCA.labs[1] <- paste("PC1 (", summary(PCA.all)$importance[2,1]*100, 
                               "%)", sep="")
PCA.labs[2] <- paste("PC2 (", summary(PCA.all)$importance[2,2]*100, 
                               "%)", sep="")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Extract PC values and combine
PCA.dat.all <- as.data.frame(PCA.all$x) %>% 
  rownames_to_column("libID") %>%
  # Select PCs
  dplyr::select(libID, PC1:PC3) %>% 
  #Add metadata
  left_join(dat.abund.norm.voom$targets)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#plots
PCA.list <- list()
#Loop through variables of interest
for(i in 1:length(vars_of_interest)){
  PCA <- PCA.dat.all %>% 

    ggplot(aes(PC1, PC2)) +
      geom_point(aes_string(color=vars_of_interest[i]),
                      size=3) +
      #Beautify
      theme_classic() +
      labs(x=PCA.labs[[1]], y=PCA.labs[[2]], 
           title="voom normalized\nabundant logCPM",
           color=vars_of_interest[i]) +
      coord_fixed(ratio=1) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))
 PCA.list[[i]] <- PCA
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=12}
plot_grid(plotlist=PCA.list, ncol=2)

dir.create("figs", showWarnings = FALSE)
ggsave("figs/PCA_SA_RSTR.Mtb_geneVoom.png", 
       plot_grid(plotlist=PCA.list, ncol=2),
       height=6, width=8)
```

### Potential co-variates

```{r echo=FALSE, warning=FALSE, message=FALSE}
#plots
PCA.list <- list()
#Loop through variables of interest
for(covar in c("age","gender","bmi","bcgscar",
               "HoursWorkedUnderground","YearsWorkedUnderground")){

  PCA <- PCA.dat.all %>% 
    arrange(desc(covar)) %>% 
    
    ggplot(aes(PC1, PC2)) +
      geom_point(aes_string(color=covar),
                      size=3) +
      #Beautify
      theme_classic() +
      labs(x=PCA.labs[[1]], y=PCA.labs[[2]], 
           title="voom normalized\nabundant logCPM",
           color=gsub("_","\n",covar)) +
      coord_fixed(ratio=1) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))
 PCA.list[[covar]] <- PCA
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=30, fig.width=8.5}
plot_grid(plotlist=PCA.list, ncol=2)

ggsave("figs/PCA_SA_RSTR.Mtb_covar.png", 
       plot_grid(plotlist=PCA.list, ncol=2),
       height=10, width=8)
```

# Interaction models 
## Linear mixed effects models 

```{r}
source("scripts/kin.model.fxn.R")
```

```{r eval=FALSE}
lmekin.loop(dat = dat.abund.norm.voom, kin = NULL, 
             x.var = c("condition","Sample_Group"), ptID="FULLIDNO",
             co.var=c("age"), 
             interaction=TRUE, 
             lm=FALSE, lme=TRUE,
             outdir="results/gene_level/", name="SA_RSTR.interaction_age",
             processors=5, p.method="BH")

lmekin.loop(dat = dat.abund.norm.voom, kin = NULL, 
             x.var = c("condition","Sample_Group"), ptID="FULLIDNO",
             interaction=TRUE, 
             lm=FALSE, lme=TRUE,
             outdir="results/model_selection/", name="SA_RSTR.interaction",
             processors=5, p.method="BH")
```

```{r message=FALSE}
model.files <- list.files(path="results/model_selection/", pattern="results.csv.gz",
                          full.names = TRUE)
#load results
gene_pval <- data.frame()

for(filename in model.files){
  gene_pval <- read_csv(filename) %>% 
    bind_rows(gene_pval)
}
```

Format model result labels

```{r}
gene_pval_all <- gene_pval %>% 
  mutate(coVar = recode_factor(factor(group), "SA_RSTR.interaction"="none",
                               "SA_RSTR.interaction_age"="age"))
```

## Model fits (sigma)
#### Age

```{r echo=FALSE}
gene_pval_all %>% 
  distinct(coVar, gene, sigma) %>% 
  pivot_wider(names_from = coVar, values_from = sigma) %>% 
  
  ggplot(aes(x=age, y=none)) +
  geom_point() +
  geom_abline(slope=1, intercept=0, color="red") +
  theme_classic() +
  labs(x="~ Mtb * RSTR + age",
       y="~ Mtb * RSTR",
       title="Model fit (sigma)")
```

```{r echo=FALSE}
gene_pval_all %>% 
  distinct(coVar, gene, sigma) %>% 
  pivot_wider(names_from = coVar, values_from = sigma) %>% 
  mutate(best = ifelse(none<age, "none", 
                       ifelse(age<none, "age", NA)),
         diff = abs(none-age)) %>% 
  group_by(best) %>% 
  summarise(n=n(),
            mean.diff = mean(diff), .groups="drop") %>% 

  kable(col.names = c("Co-variates", "Number of genes best fit", 
                      "Mean difference in sigma")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)%>% 
  collapse_rows(1, valign="top")
```

The best fit model with or without age is split by genes.

## Significant genes

```{r echo=FALSE}
fdr.cutoff <- 0.2
```

```{r echo=FALSE, fig.width=8.5, fig.height=4}
par(mfrow=c(1,2))
for(co.var in levels(gene_pval_all$coVar)){
venn.ls <- list()

pval.temp <- gene_pval_all %>% 
  filter(coVar == co.var & FDR <= fdr.cutoff)

if(nrow(pval.temp)>0){
venn.ls[["+Mtb vs media\nexpression"]] <- pval.temp %>% 
  filter(grepl("condition", variable) & !grepl("Sample_Group",variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["RSTR vs LTBI\nexpression"]] <- pval.temp %>% 
  filter(grepl("Sample_Group", variable) & !grepl("condition",variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["Interaction\nexpression"]] <- pval.temp %>% 
  filter(grepl("condition", variable) & grepl("Sample_Group", variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["Age"]] <- pval.temp %>%
  filter(grepl("age", variable)) %>%
    distinct(gene) %>% unlist(use.names = FALSE)

venn(ilab=FALSE, zcolor = "style",ilcs=1, sncs=1,
     x=venn.ls, box=FALSE)
  
  title(sub=paste(co.var, "\nFDR <", fdr.cutoff, sep=" "), 
        line = -1, cex.sub=1.5)
}
}
```

```{r echo=FALSE}
png(filename="results/model_selection/SA_model_venn_compare.png",
    width=10, height=5, units = "in", res=150)

par(mfrow=c(1,2))
for(co.var in levels(gene_pval_all$coVar)){
venn.ls <- list()

pval.temp <- gene_pval_all %>% 
  filter(coVar == co.var & FDR <= fdr.cutoff)

if(nrow(pval.temp)>0){
venn.ls[["+Mtb vs media\nexpression"]] <- pval.temp %>% 
  filter(grepl("condition", variable) & !grepl("Sample_Group",variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["RSTR vs LTBI\nexpression"]] <- pval.temp %>% 
  filter(grepl("Sample_Group", variable) & !grepl("condition",variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["Interaction\nexpression"]] <- pval.temp %>% 
  filter(grepl("condition", variable) & grepl("Sample_Group", variable)) %>% 
    distinct(gene) %>% unlist(use.names = FALSE)

venn.ls[["Age"]] <- pval.temp %>%
  filter(grepl("age", variable)) %>%
    distinct(gene) %>% unlist(use.names = FALSE)

venn(ilab=FALSE, zcolor = "style",ilcs=1, sncs=1,
     x=venn.ls, box=FALSE)
  
  title(sub=paste(co.var, "\nFDR <", fdr.cutoff, sep=" "), 
        line = -1, cex.sub=1.5)
}
}

dev.off()
```

DEGs (significant for interaction or RSTR)

```{r echo=FALSE}
temp <- gene_pval_all %>% 
  filter(grepl("Sample_Group", variable) & FDR <= fdr.cutoff) %>% 
  distinct(gene, model, coVar) %>% 
  count( model,coVar) %>% 
  rename(Total=n)

gene_pval_all %>% 
  filter(grepl("Sample_Group", variable) & FDR <= fdr.cutoff) %>% 
  mutate(variable = recode_factor(factor(variable),
                           "conditionTB:Sample_GroupRSTR"="Mtb:RSTR interaction",
                           "condition:Sample_Group"="Mtb:RSTR interaction",
                           "Sample_Group"="RSTR",
                           "Sample_GroupRSTR"="RSTR")) %>% 
  count(model, coVar, variable) %>% 
  pivot_wider(names_from = variable, values_from = n) %>% 
  mutate(RSTR = ifelse(is.na(RSTR), 0, RSTR)) %>% 
  full_join(temp) %>% 

kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  collapse_rows(1:3, valign="top") %>% 
  add_header_above(c(" "=2, "DEG FDR < 0.2"=3))
```

# R session

```{r}
sessionInfo()
```

***